The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-10 rational.tex
moved two constructions to teh section on generic constructions
constructions
negative tpes and constructions

Describing the syntax and types of an extension to {{langRev}} with additive
duality is fairly straightforward. Basically, we extend the language with
negative types denoted {{-b}}, negative values denoted {{-v}}, and two
isomorphisms {{eta+}} and {{eps+}}:

@@

The isomorphisms {{eta+}} and {{eps+}} witness the following
isomorphism:
%subcode{opsem} include main
% eta+ &: 0 <-> (-b) + b :& eps+
And as expected we have the following typing rule for negative value:
%subcode{proof} include main
%@ |- v : b
%@@ |- -v : -b

For the graphical language, we visually present {{eta+}} and {{eps+}} as
U-shaped connectors. On the left below is {{eta+}} showing the map from {{0}}
to {{-b+b}} and dually on the right is {{eps+}} showing the map from {{-b+b}}
to 0. Even though the diagrams below show a {{0}} wire for completeness,
later diagrams will always drop them in contexts where we can implicitly
introduce/eliminate then using {{zeroi}}/{{zeroe}}:
\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/eta.pdf}
\end{center}

\begin{center}
  \includegraphics{diagrams/eps.pdf}
\end{center}
\end{multicols}

The usual interpretation of the type {{b1+b2}} is that we either have an
appropriately tagged value of type {{b1}} or an appropriately tagged value of
type {{b2}}. This interpretation is maintained in the presence of {{eta+}}
and {{eps+}} in the following sense: a value of type {{right v : -b + b}}
flowing into an {{eps+}} on the lower wire switches to a value 
{{left (-v): -b + b}} that flows backwards on the upper wire. The inversion of
direction is captured by the negative sign on the value. 
As an example, the diagram representing the isomorphism from the introduction
is:
\begin{center}
  \includegraphics{diagrams/algebra1.pdf}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Compact Closure with Additive Inverses}

\paragraph*{Coherence.}
The appropriate coherence condition for Compact Closed Categories,
expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
may be visualized as explained by Selinger
\cite{springerlink:10.1007/978-3-642-12821-94} as follows:

@@

\paragraph*{Trace}

\begin{center}
  \includegraphics{diagrams/trace.pdf}
\end{center}

\paragraph*{Involution (Principium Contradictiones)}

\paragraph*{Higher-Order Functions}

\paragraph*{Function application}

\begin{center}
  \includegraphics{diagrams/apply1.pdf}
\end{center}

\begin{center}
  \includegraphics{diagrams/apply2.pdf}
\end{center}

\paragraph*{Function composition}

\begin{center}
  \includegraphics{diagrams/compose1.pdf}
\end{center}

\begin{center}
  \includegraphics{diagrams/compose.pdf}
\end{center}

This is also equivalent to sequencing both the computation blocks.

\paragraph*{Delimited continuation}

%%%%%%%%%%%%%

\paragraph*{Negative Information Flow.} 
Hence, the intuitive way to read the diagrams is that they reverse the
flow of information -- in normal {{langRev}} circuits information
flows from left-to-right and {{eps+}} operations cause information to
flow from right-to-left.  The interpretation of {{eta+}} is precisely
the opposite: it turns a right-to-left flow of information in a
circuit into a left-to-right flow.

\begin{multicols}{2}
\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eps_plus1.pdf}
}
\end{center}

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eps_plus2.pdf}
}
\end{center}  
\end{multicols}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions in {{langRevEE}} }
\label{sec:monoidal-constructions}

%%%%%%%%%%%%%%%%%%
\subsection{Constructions Specific to Negation}
\label{sec:neg-constructions}

\paragraph*{Negation distributes over {{+}}. }

\paragraph*{ {{eps_{fst} }} }

@@

\paragraph*{Lifting negation out of {{*}}. }

@@

\paragraph*{Lifting a operation of positive types to negated types}

%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions Specific to Multiplication}

\paragraph*{Distributing fractions over {{*}} }

{{1/(b1*b2) <-> (1/b1) * (1/b2)}}

\paragraph*{Multiplying Fractions}

@@

\paragraph*{Lifting Computation intro fractions}

{{b1 <-> b2 ==> 1/b1 <-> 1/b2}}

\paragraph*{Adding the numerator}

@@

\paragraph*{Fraction Addition}

>>>>

Describing the syntax and types of an extension to {{langRev}} with additive
duality is fairly straightforward. Basically, we extend the language with
negative types denoted {{-b}}, negative values denoted {{-v}}, and two
isomorphisms {{eta+}} and {{eps+}} with the following type judgements:

@@

\vspace{-15pt}
\begin{multicols}{2}  
%subcode{opsem} include main
% eta+ &: 0 <-> (-b) + b :& eps+

%subcode{proof} include main
%@ |- v : b
%@@ |- -v : -b
\end{multicols}

For the graphical language, we visually present {{eta+}} and {{eps+}} as
U-shaped connectors. On the left below is {{eta+}} showing the map from {{0}}
to {{-b+b}} and dually on the right is {{eps+}} showing the map from {{-b+b}}
to 0. Even though the diagrams below show a {{0}} wire for completeness,
later diagrams will always drop them in contexts where we can implicitly
introduce/eliminate then using {{zeroi}}/{{zeroe}}:
\begin{multicols}{2}
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/eta.pdf}
}
\end{center}

\begin{center}
\scalebox{0.8}{
 \includegraphics{diagrams/eps.pdf}
}
\end{center}
\end{multicols}

The usual interpretation of the type {{b1+b2}} is that we either have an
appropriately tagged value of type {{b1}} or an appropriately tagged value of
type {{b2}}. This interpretation is maintained in the presence of {{eta+}}
and {{eps+}} in the following sense: a value of type {{right v : -b + b}}
flowing into an {{eps+}} on the lower wire switches to a value 
{{left (-v): -b + b}} that flows backwards on the upper wire. The inversion of
direction is captured by the negative sign on the value.

\begin{multicols}{2}
\begin{center}
\scalebox{1.1}{
  \includegraphics{diagrams/eps_plus1.pdf}
}
\end{center}

\begin{center}
\scalebox{1.1}{
  \includegraphics{diagrams/eps_plus2.pdf}
}
\end{center}  
\end{multicols}

\paragraph*{Negative Information Flow.} 
Hence, the intuitive way to read the diagrams is that they reverse the
flow of information -- in normal {{langRev}} circuits information
flows from left-to-right and {{eps+}} operations cause information to
flow from right-to-left.  The interpretation of {{eta+}} is precisely
the opposite: it turns a right-to-left flow of information in a
circuit into a left-to-right flow.
As an example, the diagram representing the isomorphism from the
introduction is:
\begin{center}
  \includegraphics{diagrams/algebra1.pdf}
\end{center}

%\subsection{Compact Closure with Additive Inverses}
\paragraph*{Coherence for Compact Closure with Additive Inverses.}
The appropriate coherence condition for Compact Closed Categories,
expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
may be visualized as explained by Selinger
\cite{springerlink:10.1007/978-3-642-12821-94} as follows:

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions in the Additive Monoid}

\begin{enumerate}
\item 
\emph{Trace}

\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/trace.pdf}
\end{center}

\end{multicols}

\item
\emph{Involution (Principium Contradictiones)}

\item
\emph{Functions}

\item
\emph{Function composition.}

\begin{multicols}{2}
    \includegraphics{diagrams/compose1.pdf}

    \includegraphics{diagrams/compose.pdf}
\end{multicols}
This is also equivalent to sequencing both the computation blocks.

\item
\emph{Function application.}
test.
\begin{multicols}{2}
\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/apply1.pdf}
}
\end{center}
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/apply2.pdf}
}
\end{center}
\end{multicols}

\item
\emph{Delimited continuation}

\item
\emph{Negation distributes over {{+}}. }

\item
\emph{Lifting a operation of positive types to negated types}

@@

%% %%%%%%%%%%%%%%%%%%%%%%
%% \subsection{Constructions in {{langRevEE}} }
%% \label{sec:monoidal-constructions}

%%%%%%%%%%%%%%%%%%
\subsection{Constructions in {{langRevEE}} specific to Addition and Multiplication}
\label{sec:neg-constructions}

\begin{enumerate}

\item 
\emph{ {{eps_{fst} }} }

@@

\item 
\emph{Lifting negation out of {{*}}. }

@@

%% \emph{Distributing fractions over {{*}} }
%% {{1/(b1*b2) <-> (1/b1) * (1/b2)}}

%% \emph{Lifting Computation intro fractions}
%% {{b1 <-> b2 ==> 1/b1 <-> 1/b2}}

\item 
\emph{Multiplying Fractions}

@@

\item 
\emph{Adding the numerator}

@@

\item 
\emph{Fraction Addition}
